Lucky Thirteen - A Cryptographic Analyses

### 2013-07-10 paper.tex
Section III alpha-1

\begin{abstract}
TODO.

@@

The paper is ordered as follows. First, Section~\ref{sec:crypto} will very briefly overview the cryptography (and crypto format) that's important to understand the attack. Section~\ref{sec:TLS} describes the main features in TLS responsible for padding oracle attacks, which will be discussed in Section~\ref{sec:paddingoracle}. The wrongly placed bandits and their weaknesses are discussed in Section~\ref{sec:lucky} which covers the Lucky Thirteen attack. This paper will conclude with the future of TLS in Section~\ref{sec:future}.

@@

\subsection{Cryptographic format}
\label{sec:crypto:format}
The cryptographic format used in this paper is presented in Table~\ref{sec:crypto:format:table}.
\begin{table}[h]
    \begin{tabular}{l|l}
    Format & Meaning \\ \hline
    A $||$ B    & Concatenation of A and B  \\
    A $\oplus$ B  & The bitwise XOR of A and B     \\
    H(A)    & Hash-function applied to A    \\
    $E_k$(A)    & Encryption with key K applied to A    \\
    \end{tabular}
    \caption{Cryptographic format used in this paper}
    \label{sec:crypto:format:table}
\end{table}

\subsection{MAC - Message Authentication Code}
\label{sec:crypto:hmac}
To ensure message integrity, RFC5246~\cite{ietf2008transport} prescribes the use of a keyed Message Authentication Code (MAC). The MAC cipher suites described in the RFC are based on hashing functions, HMACs, and are described in~\cite{krawczyk1997rfc}. Though, the RFC does not prohibit the use of other cipher suites, the implementations found by~\citeauthor{alfardan2013lucky} in~\cite{alfardan2013lucky} all restrict the use to HMAC-MD5 ($16$ bytes of MAC), HMAC-SHA-1 ($20$ bytes of MAC) or HMAC-SHA-256 ($32$ bytes of MAC).

To compute the HMAC for a message M and key K, the specified HMAC algorithms use the following formula. 
\[ HMAC(K,M) = H((K \oplus opad)||H(K \oplus ipad)||M)). \]
With opad, ipad to be specific $64$-byte values and key K zero-padded to $64$ bytes. "Each hash function uses a \textit{Merkle-Damg\aa rd strengthening}. Here, an 8-byte length field followed by padding of a specified byte format are appended to the message M to be hashed. The padding is at least $1$ byte in length and aligns the data on a $64$-byte boundary. The relevant hash functions also have an iterated structure, processing messages in chunks of $64$ bytes ($512$ bits) using a compression function, with the output of each compression step being chained into the next step. The compression function in turn involves a complex round structure, with many basic arithmetic operations on data being involved in each round."~\cite{alfardan2013lucky}. This means that, depending on the size of message M, the number of runs of the hash function H can be computed. Because of the Merkle-Damg\aa rd strengthening, a message M is appended by the $8$-byte length field, and padded to be a multiple of $64$ bytes. This means that the first of the Merkle-Damg\aa rd rounds consists of $55$ bytes of message M, followed by $64$ bytes for every following $64$ bytes of message M. The inner hash function depends on the $message length of M + 1$ constant hash function round, the outer hash function is a constant of $2$. This means that the following formula represents the number of hash function calls for L is the message length of M in bytes:
\[ \lceil \frac{L - 55}{64} \rceil + 4 \]

@@

The plaintext is cut into blocks of the input size of the Block Cipher (AES has a $16$ bytes block size). Ciphertext blocks are computed as:
\[ 
\begin{split}
C_i &= E_K(P_i \oplus C_{i-1}) \\
C_0 &= IV 
\end{split}
\]
With $P_i$ the blocks of plaintext (plus extras, as described in Section~\ref{sec:TLS}) and $K$ the key for Block Cipher $E$. Ciphertext $C$ is the concatenation of all $C_i$'s.

\section{TLS}
\label{sec:TLS}
TLS 1.1 and 1.2: check the padding format carefully, report a single error message for padding and MAC failures, and make the record processing time essentially the same whether or not the padding is correct.

\begin{figure}[h]
	\centering
	\includegraphics[width=0.5\textwidth]{tls-representation.jpg}
	\caption{Representation of TLS~\cite{alfardan2013lucky}}
	\label{fig:tls}
\end{figure}

@@

\subsection{Bomb Oracles}
\label{sec:paddingoracle:bomb}
blabla $\oplus$ blabla

>>>>

\begin{abstract}
TODO: clean up the text underneath, be more elaborate, fix text from all sections.

@@

The paper is ordered as follows. First, Section~\ref{sec:crypto} will very briefly overview the cryptography (and crypto format) that's important to understand the attack. Section~\ref{sec:tls} describes the main features in TLS responsible for padding oracle attacks, which will be discussed in Section~\ref{sec:paddingoracle}. The wrongly placed bandits and their weaknesses are discussed in Section~\ref{sec:lucky} which covers the Lucky Thirteen attack. This paper will conclude with the future of TLS in Section~\ref{sec:future}.

@@

\subsection{Cryptographic format}
\label{sec:crypto:format}
The cryptographic format used in this paper is presented in Table~\ref{sec:crypto:format:table}.
\begin{table}[h]
    \begin{tabular}{l|l}
    Format & Meaning \\ \hline
    A $||$ B    & Concatenation of A and B  \\
    A $\oplus$ B  & The bitwise XOR of A and B     \\
    H(A)    & Hash-function applied to A    \\
    $E_k$(A)    & Encryption with key K applied to A    \\
    $D_k$(A)    & Decryption with key K applied to A    \\
    \end{tabular}
    \caption{Cryptographic format used in this paper}
    \label{sec:crypto:format:table}
\end{table}

\subsection{MAC - Message Authentication Code}
\label{sec:crypto:hmac}
To ensure message integrity, RFC5246~\cite{ietf2008transport} prescribes the use of a keyed Message Authentication Code (MAC). The MAC cipher suites described in the RFC are based on hashing functions, HMACs, and are described in~\cite{krawczyk1997rfc}. Though, the RFC does not prohibit the use of other cipher suites, the implementations found by~\citeauthor{alfardan2013lucky} in~\cite{alfardan2013lucky} all restrict the use to HMAC-MD5, HMAC-SHA-1 or HMAC-SHA-256~\footnote{HMAC-MD5 has 16 bytes of MAC.\\HMAC-SHA-1 has 20 bytes of MAC.\\HMAC-SHA-256 has 32 bytes of MAC.} To compute the HMAC for a message M and key K, the specified HMAC algorithms use the following formula. 
\[ HMAC(K,M) = H((K \oplus opad)||H(K \oplus ipad)||M)). \]
With \textit{opad}, \textit{ipad} to be specific $64$-byte values and key $K$ zero-padded to $64$ bytes. "Each hash function uses a \textit{Merkle-Damg\aa rd strengthening}. Here, an 8-byte length field followed by padding of a specified byte format are appended to the message M to be hashed. The padding is at least $1$ byte in length and aligns the data on a $64$-byte boundary. The relevant hash functions also have an iterated structure, processing messages in chunks of $64$ bytes ($512$ bits) using a compression function, with the output of each compression step being chained into the next step. The compression function in turn involves a complex round structure, with many basic arithmetic operations on data being involved in each round."~\cite{alfardan2013lucky}.

TODO: more elaborate; maybe a figure or a formula with regards to the above Merkle-Damg\aa rd strengthening.

This means that, depending on the size of message M, the number of runs of the hash function H can be computed. Because of the Merkle-Damg\aa rd strengthening, a message M is appended by the $8$-byte length field, and padded to be a multiple of $64$ bytes. This means that the first of the Merkle-Damg\aa rd rounds consists of $55$ bytes of message M, followed by $64$ bytes for every following $64$ bytes of message M. The inner hash function depends on the $message length of M + 1$ constant hash function round, the outer hash function is a constant of $2$. This means that the following formula represents the number of hash function calls for L is the message length of M in bytes:
\[ \lceil \frac{L - 55}{64} \rceil + 4 \]

@@

The plaintext is cut into blocks of the input size of the Block Cipher (AES has a $16$ bytes block size). Ciphertext blocks are computed as:
\[ 
\begin{split}
C_i &= E_K(P_i \oplus C_{i-1}) \\
C_0 &= IV 
\end{split}
\]
With $P_i$ the blocks of plaintext (plus extras, as described in Section~\ref{sec:tls}) and $K$ the key for Block Cipher $E$. Ciphertext $C$ is the concatenation of all $C_i$'s.

\section{TLS}
\label{sec:tls}
TLS is used to provide for a cryptographically secure connection between two entities on the internet. TLS consists of four protocols: a handshake protocol to establish a connection; an alert protocol; the change cipher spec protocol; application data protocol. In this paper, the scope will be on an already established connection between these two entities (application data protocol) rather than the other protocols because the attack by~\citeauthor{alfardan2013lucky} takes place after the handshake and without the alert of change cipher spec protocol. Furthermore, the scope is on the versions 1.1 and 1.2 of TLS. These versions have tried to fix the origin of an attack proposed by~\citeauthor{vaudenay2002security} in~\cite{vaudenay2002security}.

In the following subsections the TLS Record and the procedure of accepting a received record will be described.

\subsection{TLS Record}
\label{sec:tls:record}
The basis for the attack on TLS 1.0 has to do with the fundamental approach TLS took into the formatting their TLS Record into a MAC-Encode-Encrypt Scheme as described in Section~\ref{sec:crypto}. First the MAC is applied to the payload, preceded by a HDR~\footnote{The HDR field is a 5 byte field consisting of the version (1 byte), protocol type (2 bytes) and length of the payload.(2 bytes)} and a SQN~\footnote{The SQN field is a 8 byte field consisting of the current sequence number, which is incremented each record.} field. After which the original payload is appended by the MAC and padding, and then encrypted to become ciphertext. Figure~\ref{fig:tls} has a visual representation of an actual TLS record. Actually, this Figure contains three minor errors/simplifications: the HDR and SQN fields are switched with respect to the Figure, according to~\cite{ietf2008transport}; the padding field consists of two field, the actual padding (minimal of $1$ byte) followed by the length of this padding ($1$ byte); the ciphertext is preceded by the HDR field and the IV that is forced to be completely random and unpredictable.

\begin{figure}[h]
	\centering
	\includegraphics[width=0.5\textwidth]{tls-representation.jpg}
	\caption{Representation of TLS record.~\cite{alfardan2013lucky}}
	\label{fig:tls}
\end{figure}

\subsection{Accepting a record}
\label{sec:tls:accepting}
The RFC~\cite{ietf2008transport} does not describe the steps that need to be taken before accepting the payload of a received TLS record, fortunately the authors of the Lucky Thirteen attack have proposed a (incomplete) list of actions to be undertaken.
\begin{itemize}
  \item Check: is the ciphertext size a multiple of the block size?
  \item Check: is the ciphertext large enough to contain at least a zero-length record, a MAC tag of the required size and at least one byte of padding?
  \item Decrypt the ciphertext to recover the plaintext blocks: $P_i = D_k(C_i) \oplus C_{i-1}$.
  \item Is the padding of the appropriate format?
  \item Is the MAC correct (using the correct SEQ number and HDR field?
\end{itemize}

@@

\subsection{Bomb Oracles}
\label{sec:paddingoracle:bomb}
